---
title: 'Splicing Exploration of the Tao Dataset'
author: "Andrés Gordo Ortiz"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    number_sections: true
    df_print: kable
knit: 
  root.dir: .

editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Ensure a clean environment and load required libraries.
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE
)

# Function to install and load packages dynamically.
install_load <- function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
}

# Load required libraries
install_load("betAS")
install_load("ggplot2")
install_load("plotly")
install_load("dplyr")
install_load("cowplot")
install_load("DT")
```


```{r load inclusion table}
# Define the path to the inclusion table
path <- paste0(getwd(),"/inclusion_tables/Tao_INCLUSION_LEVELS_FULL-mm10.tab")

# Load the dataset using vast-tools functions
taodata <- getDataset(pathTables = path, tool = "vast-tools")

# Extract alternative splicing events
taoevents <- getEvents(taodata, tool = "vast-tools")

# Filter events based on PSI thresholds
taoevents <- alternativeEvents(taoevents, minPsi = 1, maxPsi = 99)

# Display summary of events per type
taoevents$EventsPerType

```

```{r load_metadata}
# Load metadata file containing sample information
metadata_path <- paste0(getwd(),"/metadata/tao_metadata.csv")
metadata <- read.csv(metadata_path, sep = ",")
DT::datatable(metadata, options = list(pageLength = 5, scrollX = TRUE))

```

```{r events extraction}
# Filter events for exons, introns, and alternative splicing events
tao_exons <- filterEvents(taoevents, types = c("C1", "C2", "C3", "S", "MIC"), N = 5)
tao_introns <- filterEvents(taoevents, types = c("IR"), N = 5)
tao_alternative <- filterEvents(taoevents, types = c("Alt3", "Alt5"), N = 5)

# Output the count of alternative splicing events
cat(paste0("Alternative splicing events detected: ", nrow(tao_exons$PSI)))

```
## Splicing Quality Check { .tabset }

### Exons
```{r bigPicturePlotExons, fig.width=8, fig.height=4}
# Create a big picture plot for exon inclusion levels
bigPicturePlotExons <- bigPicturePlot(table = tao_exons$PSI)

# Customize the plot for professional reporting
bigPicturePlotExons +
  ggtitle("Exon Inclusion Levels") +
  xlab("Samples") +
  ylab("PSI Values") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

```
### Introns
```{r bigPicturePlotIntrons, fig.width=8, fig.height=4}
# Plot for tao_exons
bigPicturePlotIntrons <- bigPicturePlot(table = tao_introns$PSI)
# Customize the plot for professional reporting
bigPicturePlotIntrons +
  ggtitle("Introns Inclusion Levels") +
  xlab("Samples") +
  ylab("PSI Values") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```
### Altenative Splicing
```{r bigPictureAlternative, fig.width=8, fig.height=4}
# Plot for tao_exons
bigPicturePlotAlternative <- bigPicturePlot(table = tao_alternative$PSI)
bigPicturePlotAlternative +
  ggtitle("Alternative Splice Site Inclusion Levels") +
  xlab("Samples") +
  ylab("PSI Values") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```

### PCA
```{r PCA calculation}
# Subset and scale data
data_pca <- taodata[, taoevents$Samples]
data_pca <- na.omit(data_pca)
pca_result <- prcomp(t(data_pca))  # Perform PCA

# Prepare PCA results for plotting
pca_data <- as.data.frame(pca_result$x)
pca_data$Sample <- rownames(pca_data)
merged_data <- merge(pca_data, metadata, by.x = "Sample", by.y = "run_accession")

```

```{r pca plot}
# Generate dynamic color palette
num_levels <- length(unique(merged_data$experiment_title))
color_palette <- RColorBrewer::brewer.pal(min(12, num_levels), "Set3")

# Create PCA plot
ggplot(merged_data, aes(x = PC1, y = PC2, color = as.factor(experiment_title))) +
  geom_point(size = 4) +
  scale_color_manual(values = color_palette) +
  labs(
    title = "PCA of Samples (PSI Data)",
    x = paste("PC1 (", round(100 * summary(pca_result)$importance[2, 1], 1), "%)", sep = ""),
    y = paste("PC2 (", round(100 * summary(pca_result)$importance[2, 2], 1), "%)", sep = "")
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank(),
    axis.title = element_text(size = 14)
  )

```


## PSI Distributions comparing 0 Days vs. 7 Days {#psi-distributions .tabset}

### Define Groups
```{r aux_featuregroup}
# Extract unique groups and sample IDs
groupingVariable <- "experiment_title"
groups <- unique(metadata[, groupingVariable])
samples <- metadata$run_accession

# Assign colors to groups
random_colors <- c("#FF9AA2", "#FFB7B2", "#FFDAC1", "#D5E1DF")

# Create group list with metadata
groupList <- lapply(1:length(groups), function(i) {
  list(
    name = groups[i],
    samples = samples[metadata[, groupingVariable] == groups[i]],
    color = random_colors[i]
  )
})
names(groupList) <- groups

```

```{r groups_auxiliary}
# Define groups
groupA    <- "0_days"
groupB    <- "7_days"
# Define samples inside each group
samplesA    <- groupList[[groupA]]$samples
samplesB    <- groupList[[groupB]]$samples

# Convert samples into indexes
colsGroupA    <- convertCols(tao_exons$PSI, samplesA)
colsGroupB    <- convertCols(tao_exons$PSI, samplesB)

set.seed(42) #Setting seed for downstream simulations of the beta distribution
```

### Exons { .tabset }

#### Pdiff
```{r plotvolcano_Pdiff, fig.width=14, fig.height=10, warning=F}
volcanoTable_Pdiff <- prepareTableVolcano(psitable = tao_exons$PSI,
                                    qualtable = tao_exons$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100, 
                                    seed=TRUE, 
                                    CoverageWeight = FALSE)

volcano_Pdiff <- plotVolcano(betasTable = filter(volcanoTable_Pdiff,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_Pdiff + labs(title = "Volcano Plot: Pdiff")
 
```

```{r Pdiff_table}
exon_diff_tab <- volcanoTable_Pdiff[, c("GENE", "EVENT", "COORD", "Pdiff", "deltapsi")] %>%
  filter(!is.na(EVENT), Pdiff >= 0.95) %>%  # Remove rows where EVENT is NA and Pdiff >= 0.95
  arrange(desc(Pdiff), desc(abs(deltapsi)))  # Sort by Pdiff and abs(deltapsi)
rownames(exon_diff_tab) <- 1:nrow(exon_diff_tab)

# Render DataTable with enhancements
datatable(
  exon_diff_tab,
  options = list(
    pageLength = 10,                      # Rows per page
    autoWidth = TRUE,                     # Adjust column widths automatically
    dom = 'Bfrtip',                       # Add buttons for export
    buttons = c("copy", "csv", "excel", "pdf"),  # Simplified button definitions
    columnDefs = list(
      list(targets = "_all", className = "dt-center"),    # Center-align all columns
      list(
        targets = 3,  # Highlight significant Pdiff
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && parseFloat(data) > 0.99) {
               return '<span style=\"color: green; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      ),
      list(
        targets = 4,  # Highlight large deltapsi values
        render = JS(
          "function(data, type, row) {
             if (type === 'display' && Math.abs(parseFloat(data)) > 0.5) {
               return '<span style=\"color: red; font-weight: bold;\">' + data + '</span>';
             }
             return data;
           }"
        )
      )
    )
  ),
  rownames = FALSE,                       # Disable row names
  extensions = "Buttons"                  # Enable export buttons
) %>%
  formatStyle(
    columns = "Pdiff",
    color = styleInterval(0.99, c("black", "green")),
    fontWeight = styleEqual(0.99, "bold")  # Ensure consistent styling
  ) %>%
  formatStyle(
  columns = "deltapsi",
  color = styleInterval(c(-0.5, 0.5), c("black", "red", "black")),  # Three colors for two intervals
  fontWeight = styleInterval(c(-0.5, 0.5), c("normal", "bold", "normal"))  # Three weights for two intervals
)


```

#### F-statistic


```{r plotvolcano_Fstat, fig.width=14, fig.height=10, warning=F}
volcanoTable_Fstat <- prepareTableVolcanoFstat(psitable = tao_exons$PSI,
                                    qualtable = tao_exons$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100, 
                                    seed=TRUE,
                                   CoverageWeight=F )

volcano_Fstat <- plotVolcanoFstat(betasTable = filter(volcanoTable_Fstat,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_Fstat
 
```

```{r Fstat_table}
exon_fstat_tab <- volcanoTable_Fstat[,c("GENE","EVENT","COORD","Fstat","deltapsi")] %>%
  filter(!is.na(EVENT)) %>%  # Remove rows where EVENT is NA
  arrange(desc(Fstat), desc(abs(deltapsi)))  # Sort by Pdiff and abs(deltapsi)
rownames(exon_fstat_tab)<-1:nrow(exon_fstat_tab)

datatable(exon_fstat_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))
```

#### False positive rate (FPR)

To calculate the **FPR** for all events in our dataset, we will use the
`prepareTableVolcanoFDR` function. To visualize the results in a volcano
plot, we will use the `plotVolcanoFDR` function. The user can also use
the approach presented above to make the plot interactive.

```{r plotvolcano_FDR, fig.width=14, fig.height=10, warning=F}
volcanoTable_FDR <- prepareTableVolcanoFDR(psitable = tao_exons$PSI,
                                    qualtable = tao_exons$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100,
                                    nsim = 100, 
                                    seed=TRUE, 
                                    CoverageWeight = FALSE) 
        
volcano_FDR <- plotVolcanoFDR(betasTable = filter(volcanoTable_FDR,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_FDR
```

```{r FDR_table}
exon_fdr_tab <- volcanoTable_FDR[,c("GENE","EVENT","COORD","FDR","deltapsi")] %>%
  filter(!is.na(EVENT), FDR<=0.05) %>%  # Remove rows where EVENT is NA
  arrange((FDR), desc(abs(deltapsi)))  # Sort by Pdiff and abs(deltapsi)
rownames(exon_fdr_tab)<-1:nrow(exon_fdr_tab)

datatable(exon_fdr_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))
```

#### Combination Table
```{r combination table exons}
# Find common EVENTs across all tables
common_events <- Reduce(intersect, list(exon_diff_tab$EVENT, exon_fstat_tab$EVENT, exon_fdr_tab$EVENT))

# Filter each table to include only rows with common EVENTs
filtered_exon_diff_tab <- exon_diff_tab %>% filter(EVENT %in% common_events)
filtered_exon_fstat_tab <- exon_fstat_tab %>% filter(EVENT %in% common_events)
filtered_exon_fdr_tab <- exon_fdr_tab %>% filter(EVENT %in% common_events)


# Select only unique columns from each table
filtered_exon_diff_tab <- filtered_exon_diff_tab %>% select(GENE,EVENT,COORD, deltapsi,Pdiff) # Extra column is Pdiff
filtered_exon_fstat_tab <- filtered_exon_fstat_tab %>% select(EVENT, Fstat) # Extra column is Fstat
filtered_exon_fdr_tab <- filtered_exon_fdr_tab %>% select(EVENT, FDR) # Extra column is FDR

# Merge the data frames by EVENT
exon_combination_tab <- filtered_exon_diff_tab %>%
  inner_join(filtered_exon_fstat_tab, by = "EVENT") %>%
  inner_join(filtered_exon_fdr_tab, by = "EVENT") %>%
  distinct() %>%
  arrange(desc(abs(deltapsi)), desc(Fstat),desc(FDR), desc(Pdiff))

datatable(exon_combination_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))))  # Center align columns

output_dir <- file.path(getwd(), "results", "tables")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(exon_combination_tab, 
          file = file.path(output_dir, paste0("tao_exon_combination_table_", format(Sys.Date(), "%Y%m%d"), ".csv")),
          row.names = FALSE)

```


### Introns { .tabset }

#### Pdiff
```{r plotvolcano_Pdiff introns, fig.width=14, fig.height=10, warning=F}
volcanoTable_Pdiff <- prepareTableVolcano(psitable = tao_introns$PSI,
                                    qualtable = tao_introns$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100, 
                                    seed=TRUE, 
                                    CoverageWeight = FALSE)

volcano_Pdiff <- plotVolcano(betasTable = filter(volcanoTable_Pdiff,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_Pdiff
 
```

```{r Pdiff_table introns}
intron_diff_tab <- volcanoTable_Pdiff[, c("GENE", "EVENT", "COORD", "Pdiff", "deltapsi")] %>%
  filter(!is.na(EVENT), Pdiff>=0.95) %>%  # Remove rows where EVENT is NA and Pdiff greater than 0.95
  arrange(desc(Pdiff), desc(abs(deltapsi)))  # Sort by Pdiff and abs(deltapsi)
rownames(intron_diff_tab)<-1:nrow(intron_diff_tab)

datatable(intron_diff_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))
```

#### F-statistic


```{r plotvolcano_Fstat introns, fig.width=14, fig.height=10, warning=F}
volcanoTable_Fstat <- prepareTableVolcanoFstat(psitable = tao_introns$PSI,
                                    qualtable = tao_introns$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100, 
                                    seed=TRUE,
                                   CoverageWeight=F )

intron_fstat_tab <- volcanoTable_Fstat[,c("GENE","EVENT","COORD","Fstat","deltapsi")] %>%
  filter(!is.na(EVENT)) %>%  # Remove rows where EVENT is NA
  arrange(desc(Fstat), desc(abs(deltapsi)))  # Sort by Pdiff and abs(deltapsi)
rownames(intron_fstat_tab)<-1:nrow(intron_fstat_tab)

volcano_Fstat <- plotVolcanoFstat(betasTable = filter(intron_fstat_tab,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_Fstat
```

```{r Fstat_table introns}

datatable(intron_fstat_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))
```

#### False positive rate (FPR)

To calculate the **FPR** for all events in our dataset, we will use the
`prepareTableVolcanoFDR` function. To visualize the results in a volcano
plot, we will use the `plotVolcanoFDR` function. The user can also use
the approach presented above to make the plot interactive.

```{r plotvolcano_FDR introns, fig.width=14, fig.height=10, warning=F}
volcanoTable_FDR <- prepareTableVolcanoFDR(psitable = tao_introns$PSI,
                                    qualtable = tao_introns$Qual,
                                    npoints = 500,
                                    colsA = colsGroupA,
                                    colsB = colsGroupB,
                                    labA = groupA,
                                    labB = groupB,
                                    basalColor = "#89C0AE",
                                    interestColor = "#E69A9C",
                                    maxDevTable = maxDevSimulationN100,
                                    nsim = 100, 
                                    seed=TRUE, 
                                    CoverageWeight = FALSE) 


volcano_FDR <- plotVolcanoFDR(betasTable =filter(volcanoTable_FDR,!is.na(EVENT)),
                            labA = groupA,
                            labB = groupB,
                            basalColor = "#89C0AE",
                            interestColor = "#E69A9C") 

volcano_FDR
```

```{r FDR_table introns}
intron_fdr_tab <- volcanoTable_FDR[,c("GENE","EVENT","COORD","FDR","deltapsi")] %>%
  filter(!is.na(EVENT), FDR<=0.05) %>%  # Remove rows where EVENT is NA
  arrange((FDR), desc(abs(deltapsi)))  # Sort by FDR and abs(deltapsi)
rownames(intron_fdr_tab)<-1:nrow(intron_fdr_tab) 

datatable(intron_fdr_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))
```

#### Combination Table
```{r combination table introns}
# Find common EVENTs across all tables
common_events <- Reduce(intersect, list(intron_diff_tab$EVENT, intron_fstat_tab$EVENT, intron_fdr_tab$EVENT))

# Filter each table to include only rows with common EVENTs
filtered_intron_diff_tab <- intron_diff_tab %>% filter(EVENT %in% common_events)
filtered_intron_fstat_tab <- intron_fstat_tab %>% filter(EVENT %in% common_events)
filtered_intron_fdr_tab <- intron_fdr_tab %>% filter(EVENT %in% common_events)

# Select only unique columns from each table
filtered_intron_diff_tab <- filtered_intron_diff_tab %>% select(GENE,EVENT,COORD, deltapsi,Pdiff) # Extra column is Pdiff
filtered_intron_fstat_tab <- filtered_intron_fstat_tab %>% select(EVENT, Fstat) # Extra column is Fstat
filtered_intron_fdr_tab <- filtered_intron_fdr_tab %>% select(EVENT, FDR) # Extra column is FDR

# Merge the data frames by EVENT
intron_combination_tab <- filtered_intron_diff_tab %>%
  inner_join(filtered_intron_fstat_tab, by = "EVENT") %>%
  inner_join(filtered_intron_fdr_tab, by = "EVENT") %>%
  distinct() %>%
  arrange(desc(abs(deltapsi)), desc(Fstat),desc(FDR), desc(Pdiff))

datatable(intron_combination_tab, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))))  # Center align columns

# Output CSV table in the results/table folder, with todays name on it to know when it was downloaded from this script
output_dir <- file.path(getwd(), "results", "tables")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(intron_combination_tab, 
          file = file.path(output_dir, paste0("tao_intron_combination_table_", format(Sys.Date(), "%Y%m%d"), ".csv")),
          row.names = FALSE)
```

### Event Densities
```{r densities_per_event, fig.width=8, fig.height=4, warning=F}
tdensities <- plotIndividualDensitiesList(eventID = "MmuINT1026428",
                                          npoints = 500,
                                          psitable = tao_introns$PSI,
                                          qualtable = tao_introns$Qual,
                                          groupList = groupList,
                                          maxDevTable = maxDevSimulationN100, 
                                          seed=TRUE, 
                                          CoverageWeight = FALSE)

tdensities + theme_minimal() + ggtitle("MmuINT1026428")

```

```{r densities_per_event 2, fig.width=8, fig.height=4, warning=F}
eventID<-"MmuEX0042889"
tdensities <- plotIndividualDensitiesList(eventID =eventID,
                                          npoints = 500,
                                          psitable = tao_exons$PSI,
                                          qualtable = tao_exons$Qual,
                                          groupList = groupList,
                                          maxDevTable = maxDevSimulationN100, 
                                          seed=TRUE, 
                                          CoverageWeight = FALSE)

tdensities + theme_minimal() + ggtitle(eventID)
```


```{r resultsplots_per_event, fig.width=14, fig.height=4, warning=F}

plotPdiff <- prepareTableEvent(eventID = eventID,
                               psitable = tao_exons$PSI,
                               qualtable = tao_exons$Qual,
                               npoints = 500,
                               colsA = colsGroupA,
                               colsB = colsGroupB,
                               labA = groupA,
                               labB = groupB,
                               basalColor = "#89C0AE",
                               interestColor = "#E69A9C",
                               maxDevTable = maxDevSimulationN100,
                               nsim = 1000, 
                               seed=T,
                                CoverageWeight=F) %>% 
  plotPDiffFromEventObjList()

plotFstat <- prepareTableEvent(eventID = eventID,
                               psitable = tao_exons$PSI,
                               qualtable = tao_exons$Qual,
                               npoints = 500,
                               colsA = colsGroupA,
                               colsB = colsGroupB,
                               labA = groupA,
                               labB = groupB,
                               basalColor = "#89C0AE",
                               interestColor = "#E69A9C",
                               maxDevTable = maxDevSimulationN100,
                               nsim = 1000,
                               seed=T,
                             CoverageWeight=F) %>% 
  plotFstatFromEventObjList()
 
plotFDR <- prepareTableEvent(eventID = eventID,
                               psitable = tao_exons$PSI,
                               qualtable = tao_exons$Qual,
                               npoints = 500,
                               colsA = colsGroupA,
                               colsB = colsGroupB,
                               labA = groupA,
                               labB = groupB,
                               basalColor = "#89C0AE",
                               interestColor = "#E69A9C",
                               maxDevTable = maxDevSimulationN100,
                               nsim = 1000,
                               seed=T,
                             CoverageWeight=F) %>% 
  plotFDRFromEventObjList()

plot_grid(plotPdiff,plotFstat,plotFDR, ncol=3)

```

## ANOVA Multigroups

### Exons { .tabset}
```{r table_multgroups exons}
volcanoTable_multgroups_exons <- prepareTableVolcanoMultipleGroups(psitable = tao_exons$PSI,
                                                 qualtable = tao_exons$Qual,
                                                 groupList = groupList,
                                                 npoints = 500,
                                                 maxDevTable = maxDevSimulationN100, seed=TRUE, CoverageWeight = F)


```

```{r cat table multgroups}
multgroups_tab_exons <- volcanoTable_multgroups_exons[,c("GENE","EVENT","Fstat","Pdiff","Pzero","medianBetweens","deltaAbsolute")] %>%
  filter(!is.na(EVENT), Pdiff>=0.95) %>%  # Remove rows where EVENT is NA and Pdiff greater than 0.95
  arrange(desc(Fstat),desc(Pdiff))  # Sort by FDR and abs(deltapsi)
rownames(multgroups_tab_exons)<-1:nrow(multgroups_tab_exons) 

datatable(multgroups_tab_exons, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))

output_dir <- file.path(getwd(), "results", "tables")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(multgroups_tab_exons, 
          file = file.path(output_dir, paste0("tao_exon_multgroup_tab", format(Sys.Date(), "%Y%m%d"), ".csv")),
          row.names = FALSE)
```

### Introns { .tabset}
```{r table_multgroups introns}
volcanoTable_multgroups_introns <- prepareTableVolcanoMultipleGroups(psitable = tao_introns$PSI,
                                                 qualtable = tao_introns$Qual,
                                                 groupList = groupList,
                                                 npoints = 500,
                                                 maxDevTable = maxDevSimulationN100, seed=TRUE, CoverageWeight = F)


```

```{r cat table multgroups introns}
multgroups_tab_introns <- volcanoTable_multgroups_introns[,c("GENE","EVENT","Fstat","Pdiff","Pzero","medianBetweens","deltaAbsolute")] %>%
  filter(!is.na(EVENT), Pdiff>=0.95) %>%  # Remove rows where EVENT is NA and Pdiff greater than 0.95
  arrange(desc(Fstat),desc(Pdiff))  # Sort by FDR and abs(deltapsi)
rownames(multgroups_tab)<-1:nrow(multgroups_tab) 

datatable(multgroups_tab_introns, options = list(
  pageLength = 10,  # Number of rows to show per page
  autoWidth = TRUE,  # Auto adjust column widths
  dom = 'Bfrtip',  # Add buttons (for download, etc.)
  buttons = c('copy', 'csv', 'excel'),  # Enable export options
  columnDefs = list(list(targets = 0:4, className = 'dt-center'))  # Center align columns
))

output_dir <- file.path(getwd(), "results", "tables")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(multgroups_tab_introns, 
          file = file.path(output_dir, paste0("tao_introns_multgroup_tab", format(Sys.Date(), "%Y%m%d"), ".csv")),
          row.names = FALSE)
```

The multigroup ANOVA testing seems to be more restrictive in finding alternative splicing events compared to a pairwise comparison:

Multigroup ANOVA:

Compares means (or distributions, in the case of betAS) across multiple groups simultaneously.
It accounts for variability both within groups and between groups, making it more conservative.
Requires a stricter threshold to claim a significant effect due to the higher degrees of freedom and the need to correct for multiple comparisons.
Results are less likely to detect small differences specific to one or two groups.
Pairwise Comparisons:

Focuses on comparing two groups at a time.
Is more sensitive to differences between those two specific groups.
Does not account for variation across other groups, potentially identifying more splicing events as significant.
If your goal is to find more splicing events, pairwise comparisons are less restrictive, but they come at the cost of a higher false discovery rate. Use multigroup ANOVA when you want a stricter analysis and need to account for variations across all groups simultaneously.